Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-08 linear_adia.tex

\subsection{An upper limit to the cooling time}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta_c$ that allows growth of the VSI. In Appendix \ref{disp_neut_limit}
we derive the following relation between $\beta_c$ and the wave
frequency $\hat{\omega}_c$, \emph{assuming} marginal stability to the
VSI:  
\begin{subequations}\begin{align}
    \left(\hat{\omega}_c\khat\right)^4  =& \left(\hat{\omega}_c\khat\right)^2 
  + M(M+1)\left(1-h^2q^2\khat^2\right) ,\label{relax_cond_simp1}  \\
  \left(\hat{\omega}_c\khat\right)^2 = & \frac{\beta_c}{h q} (\gamma-1)(1+2M)^2
  \left(\hat{\omega}_c\khat\right) \label{relax_cond_simp2}\\
  &- 2M(M+1) . \notag
\end{align}\end{subequations}
These equations consider $\khat^2\gg 1$ and %, $M\lesssim O(1)$; 
assumes $|\hat{\omega}\khat| $ is $O(1)$ and $\beta_c\ll 1$.   Note that 
Eq. \ref{relax_cond_simp1} reduces to the  
dispersion relation for low-frequency inertial waves in the absence of
vertical shear (see Appendix \ref{stable_novshear} for details). 
%marginal stability requires hqk \leq 1

We are most interested in the longest cooling time which allows growth
for any $M$. In Appendix \ref{max_cool}, we find that if
$|hq\khat|\leq 1$, then $\beta_c$ decreases with increasing $M$.   
%\begin{align}
%  |hq\khat|^2 \leq 1 \Rightarrow \frac{d\beta_c}{dM} < 0 \quad \forall
%  M\geq 0. \label{bcrit_gen}
%\end{align}
%Eq. \ref{bcrit_gen} indicates that higher $M$ modes have lower $\beta_c$,
%i.e.\ require faster cooling for VSI growth. This trend is confirmed
%numerically in \S\ref{therm_relax_eff}.  
The VSI criterion is then set by $\beta_c$ for the
fundamental mode ($M = 0$), for which the exact solution to 
Eq. \ref{relax_cond_simp1} --- \ref{relax_cond_simp2} is
$\hat{\omega}_c\khat = \beta_c(\gamma-1)/hq = -1$. We thus find the
cooling criterion for VSI growth is    
 \begin{align}\label{iso_vsi_cond}
   \beta <   \beta_\mathrm{crit}  \equiv
   \frac{h|q|}{\gamma-1} . 
 \end{align}
 The thin disk approximation, $h \ll 1$, indicates that $\beta_\mathrm{crit} \ll
 1$, as assumed to obtain Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2}. 
 We heuristically explain $\beta_\mathrm{crit}$ in \S\ref{vsi_require}, and numerically
 test its validity in \S\ref{bcrit_num_test}.

>>>>

\subsection{The critical cooling time derived}\label{iso_vsi_beta_crit}
Here we estimate the maximum thermal relaxation timescale 
$\beta_c$ that allows growth of the VSI. In Appendix \ref{disp_neut_limit}
we derive the following relation between $\beta_c$ and the wave
frequency $\hat{\omega}_c$, \emph{assuming} marginal stability to the
VSI:  
\begin{subequations}\begin{align}
    \left(\hat{\omega}_c\khat\right)^4  =& \left(\hat{\omega}_c\khat\right)^2 
  + M(M+1)\left(1-h^2q^2\khat^2\right) ,\label{relax_cond_simp1}  \\
  \left(\hat{\omega}_c\khat\right)^2 = & \frac{\beta_c}{h q} (\gamma-1)(1+2M)^2
  \left(\hat{\omega}_c\khat\right) \label{relax_cond_simp2}\\
  &- 2M(M+1) . \notag
\end{align}\end{subequations}
These equations consider $\khat^2\gg 1$ and %, $M\lesssim O(1)$; 
assumes $|\hat{\omega}\khat| $ is $O(1)$ and $\beta_c\ll 1$.   Note that 
Eq. \ref{relax_cond_simp1} reduces to the  
dispersion relation for low-frequency inertial waves in the absence of
vertical shear (see Appendix \ref{stable_novshear} for details). 
%marginal stability requires hqk \leq 1

We are most interested in the longest cooling time which allows growth
for any $M$. In Appendix \ref{max_cool}, we find that if
$|hq\khat|\leq 1$, so that marginal stability exists, then $\beta_c$
decreases with increasing $M$.    
%\begin{align}
%  |hq\khat|^2 \leq 1 \Rightarrow \frac{d\beta_c}{dM} < 0 \quad \forall
%  M\geq 0. \label{bcrit_gen}
%\end{align}
%Eq. \ref{bcrit_gen} indicates that higher $M$ modes have lower $\beta_c$,
%i.e.\ require faster cooling for VSI growth. This trend is confirmed
%numerically in \S\ref{therm_relax_eff}.  
The VSI criterion is then set by $\beta_c$ for the
fundamental mode ($M = 0$), for which the exact solution to 
Eq. \ref{relax_cond_simp1} --- \ref{relax_cond_simp2} is
$\hat{\omega}_c\khat = \beta_c(\gamma-1)/hq = -1$. We thus find the
cooling criterion for VSI growth is    
 \begin{align}\label{iso_vsi_cond}
   \beta <   \beta_\mathrm{crit}  \equiv
   \frac{h|q|}{\gamma-1} . 
 \end{align}
 The thin disk approximation, $h \ll 1$, indicates that $\beta_\mathrm{crit} \ll
 1$, as assumed to obtain Eq. \ref{relax_cond_simp1}---\ref{relax_cond_simp2}. 
 We heuristically explain $\beta_\mathrm{crit}$ in \S\ref{vsi_require}, and numerically
 test its validity in \S\ref{bcrit_num_test}.

### 2015-05-08 paper.tex

We are more interested, however, the the global instability criterion for disturbances at all heights.
Our desired result 
is the cooling requirement 
\begin{align}\label{prelim_bcrit}
  t_c <  h |q| \OmK^{-1}/ (\gamma - 1)
\end{align}   
for VSI growth that we derive in \S\ref{iso_vsi_beta_crit}.  We can simply 
(but non-rigourously) obtain Eq. \ref{prelim_bcrit}
  by evaulating  Eq. \ref{bcrit_est} at $|z|=\gamma H/2$. %i.e., at $z=O(H)$

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.7cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx10}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.46cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx50}
  \caption{Estimate of the thermal timescale  in a fiducial 
    PPD model, the Minimum Mass Solar Nebula. The red
    line corresponds to a characteristic thermal timescale, derived in
    this work, below which the VSI is expected to operate effectively
    (the region outside the line).    
    \label{bcrit_mmsn2d} 
  }
\end{figure}

@@

% Since cooling by radiative diffusion depends on a perturbation's lengthscale, we
% plot $t_c$ for two values of the radial wavenumber,  $k_x$,  the relevant 
% component for VSI modes that are radially narrow.  The red line
%marks the boundary implied by Eq. \ref{prelim_bcrit}.  We caution that 
% Eq. \ref{prelim_bcrit} applies when $t_c$ is vertically uniform.  This section is 
% aimed at basic insights; while more realistic spatial variations in
% $t_c$ are considered in \S\ref{application}. 
% %analyzes the VSI with realistic spatial variations in.
% 
% Due to the smaller values of $t_c$ in the outer disk, we generally expect the VSI to 
% operate outside $\sim 10$ AU.  This conclusion changes however, if
% the dust opacity is reduced,  
%slowing down cooling, as we show in \S\ref{application}. % (e.g.\ Fig.\
%% \ref{mmsn_bcrit_bcool}). 
%In the inner disk, thermal times are long in the midplane, but short at higher altitutdes. 
%Again, this structure necessitates the more detailed analyses that follow.  Nevertheless, we can already 
%see that due to radiative diffusion higher wavenumbers can cool more rapidly.  Thus smaller scale disturbances 
%may trigger VSI at smaller disk radii. 
%
%On the other hand, one cannot obtain arbitrarily short cooling times
%by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
%\citetalias{barker15}), because very small wavelengths are in the
%optically thin cooling regime,  where cooling is independent of
%lengthscale. Furthermore, 
%%while we do not include viscosity in our
%%idealized disk models, in reality 
%small-scale perturbations will be 
%prone to viscous damping, including the effective viscosity from turbulent motions.
%While our linear, inviscid calculations do not address this issue directly, we expect that 
%shorter wavelength modes will drive weaker non-linear (and non-axisymmetric) transport. 
%
%% However, we cannot and do not find that VSI will always be relevant for
%% sufficiently small perturbations because \begin{inparaenum}[(i)] \item very small wavelengths are in the optically 
%% thin cooling regime,  where cooling is independent of lengthscale;
%% and \item small scale perturbations 
%% are more prone to viscous damping, and should drive weaker transport.
%% \end{inparaenum} 
%
%We now proceed with our detailed analysis of the VSI to arrive
%at Eq. \ref{prelim_bcrit} and verify it numerically. The interested 
%reader may jump to \S\ref{application} for the application of our
%results to PPDs.

>>>>

We are more interested, however, the the global instability criterion for disturbances at all heights.
Our desired result 
is the cooling requirement 
\begin{align}\label{prelim_bcrit}
  t_c <  &\underbrace{\frac{h |q| }{\gamma - 1}}_{\beta_\mathrm{crit}}\OmK^{-1}
 % &\equiv \beta_\mathrm{crit}\OmK^{-1},\notag
\end{align}   
for VSI growth that we derive in \S\ref{iso_vsi_beta_crit}.  We can simply 
(but non-rigourously) obtain Eq. \ref{prelim_bcrit}
  by evaulating  Eq. \ref{bcrit_est} at $|z|=\gamma H/2$. %i.e., at $z=O(H)$

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.7cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx10}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.46cm 0cm
  0.73cm]{figures/bcrit_mmsn2d_kx50}
  \caption{Estimate of the thermal timescale  in a fiducial 
    PPD model, the Minimum Mass Solar Nebula. The red
    lines corresponds to a maximum thermal timescale, derived in
    this work, below which the VSI is expected to operate effectively
    (the region bounded by the red lines).     
    \label{bcrit_mmsn2d} 
  }
\end{figure}

@@

% Since cooling by radiative diffusion depends on a perturbation's lengthscale, we
% plot $t_c$ for two values of the radial wavenumber,  $k_x$,  the relevant 
% component for VSI modes that are radially narrow.  The red line
%marks the boundary implied by Eq. \ref{prelim_bcrit}.  We caution that 
% Eq. \ref{prelim_bcrit} applies when $t_c$ is vertically uniform.  This section is 
% aimed at basic insights; while more realistic spatial variations in
% $t_c$ are considered in \S\ref{application}. 
% %analyzes the VSI with realistic spatial variations in.
% 
% Due to the smaller values of $t_c$ in the outer disk, we generally expect the VSI to 
% operate outside $\sim 10$ AU.  This conclusion changes however, if
% the dust opacity is reduced,  
%slowing down cooling, as we show in \S\ref{application}. % (e.g.\ Fig.\
%% \ref{mmsn_bcrit_bcool}). 
%In the inner disk, thermal times are long in the midplane, but short at higher altitutdes. 
%Again, this structure necessitates the more detailed analyses that follow.  Nevertheless, we can already 
%see that due to radiative diffusion higher wavenumbers can cool more rapidly.  Thus smaller scale disturbances 
%may trigger VSI at smaller disk radii. 
%
%On the other hand, one cannot obtain arbitrarily short cooling times
%by invoking arbitrarily small scales (cf. \citealt{urpin03}, 
%\citetalias{barker15}), because very small wavelengths are in the
%optically thin cooling regime,  where cooling is independent of
%lengthscale. Furthermore, 
%%while we do not include viscosity in our
%%idealized disk models, in reality 
%small-scale perturbations will be 
%prone to viscous damping, including the effective viscosity from turbulent motions.
%While our linear, inviscid calculations do not address this issue directly, we expect that 
%shorter wavelength modes will drive weaker non-linear (and non-axisymmetric) transport. 
%
%% However, we cannot and do not find that VSI will always be relevant for
%% sufficiently small perturbations because \begin{inparaenum}[(i)] \item very small wavelengths are in the optically 
%% thin cooling regime,  where cooling is independent of lengthscale;
%% and \item small scale perturbations 
%% are more prone to viscous damping, and should drive weaker transport.
%% \end{inparaenum} 
%
%We now proceed with our detailed analysis of the VSI to arrive
%at Eq. \ref{prelim_bcrit} and verify it numerically. The interested 
%reader may jump to \S\ref{application} for the application of our
%results to PPDs.     
>>>>>>> 3c00bb4e2b076179e280f2c07728086a110764ef
